An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-09-30 nips_2016.tex
Touch-ups and confirmed the validity of Poisson-Multinomial stuff

We start with a simple Gaussian mixture model, borrowing an experiment
from~\cite{langevin_2011}.  The parameter is 2-D, $\theta =
(\theta_1,\theta_2)$, and the parameter/data generation process is
\begin{equation}\label{eq:data_generation}
    (\theta_1,\theta_2) \sim \mathcal{N}((0,0), {\rm diag}(\sigma_1^2, \sigma_2^2));\quad \quad x_i \sim
    0.5 \cdot \mathcal{N}(\theta_1, \sigma_x^2) + 0.5 \cdot \mathcal{N}(\theta_1+\theta_2, \sigma_x^2).
\end{equation}
We set $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$.  We fix $\theta =
(0,1)$. The original paper sampled 100 data points form this distribution,
and estimated the posterior in the parameters. We are interested in performance
on larger problems and so sampled 1,000,000 points. The posterior
$p(\theta)\prod_{i=1}^{1,000,000}p(x_i\mid \theta)$, with the prior based on the
$\theta$ generation process in Equation~\ref{eq:data_generation}. The larger
number of samples produces a much sharper posterior with two very narrow peaks.
Our goal is to reproduce the original posterior, so we adjust the temperature
to $T=10,000$. Taking logs, we get the target as shown in the far left of
Figure~\ref{fig:gauss_mix_1}.

We run MCMC with our MH test using minibatch size $m=100$.  We also
run this using the methods from~\cite{cutting_mh_2014} (``Korattikara 2014'')
and~\cite{icml2014c1_bardenet14} (``Bardenet 2014'').
For~\cite{cutting_mh_2014}, we use $m=100$ and increment minibatches by 100 as
recommended for that test, and for~\cite{icml2014c1_bardenet14}, we use $m=100$
and increase the minibatch size geometrically with a ratio of $\gamma = 1.5$ as
recommended. These tests use different minibatch size updates because of their
different approaches to repeated testing. \cite{icml2014c1_bardenet14} uses
straightforward discounting, allocating a fraction of the test error $\delta$ to
each minibatch test (call these subtests). Its important to bound the number of
subtests therefore, so that the tolerable error per subtest is not too small.
\cite{cutting_mh_2014} uses a more aggressive repeated testing approach,
explicitly modeling correlation between tests and allowing many more tests for a
given $\delta$.

The tolerance for making a decision in~\cite{cutting_mh_2014} is
$\epsilon=0.005$, and the error bound control parameters
in~\cite{icml2014c1_bardenet14} are $p = 2$ and $\delta = 0.01$.  All methods
(including ours) use the same random walk proposer with covariance $\Sigma={\rm
diag}(0.3, 0.3)$.  Since the proposer generates a random walk, all shaping of
distribution of the samples is due to the M-H test. All methods are run 5000
times to collect 5000 samples.

@@

Figure~\ref{fig:gauss_mix_1} shows scatter plots of the resulting $\theta$
samples for the three methods, with darker regions indicating a greater density
of points. There are no obvious differences in these distributions. To
better understand any differences, we measure similarity between each set
of samples and the actual posterior.

To effect the comparison, we first discretize the posterior
coordinates into bins (non-overlapping ranges) in $\theta_1$ and
$\theta_2$. The probability of a sample falling into one of these bins
is the integral of the true posterior probability over the area of
that bin. Let this probability be denoted $P_i$ where $i$ is the bin
number. A single sample from any of the M-H methods should therefore
be multinomial with distribution $P$, and $n$ samples (which should
ideally be independent) should follow $\sim {\rm Multinomial}(P,n)$.
Since the ideal distribution is simple, we can use it to measure the
quality of the sample distributions rather than general purpose tests
like KL-divergence or likelihood-ratio.  The latter in any case are
problematic with zero counts in some bins as is the case here.

In practice its difficult to compute multinomial probabilities for
large $n$, and in that case the per-bin distributions are
well-approximated by Poisson distributions with parameter
$\lambda_i=P_i n$ in each bin. Given a set of samples $X_i$, let
$c_j$ denote the number of samples that fall in bin $j$:
$$c_j = |\{j : Bin(X_i) = j\}|$$
Then the log probability for the sample is
\begin{equation}
  \sum_{j=1}^{N_{\rm bins}} c_j \log (n P_j) - n P_j - \log(\Gamma(c_j+1))
\end{equation}

@@

\begin{table}[t]
    \caption{{\color{blue} Daniel: TODO Poisson Probabilities}}
    \label{tab:poissons}
    \centering
    \begin{tabular}{l l l l}
    \toprule
    Model & Ours & Korattikara 2014 & Bardenet 2014 \\
    \midrule
    Poission Value & TODO & TODO & TODO \\
    \bottomrule
    \end{tabular}
\end{table}

While these results provide general guidance on the relative accuracy of the models
its difficult interpret the quantitative scores. We can also perform significance tests to
test the null hypothesis that the ideal distribution really generated the samples.
We performed two tests:
\begin{enumerate}
  \item An exact test using simulation to generate a (approximate) distribution
    of scores given the null hypothesis.
  \item An approximate test using the chi-squared distribution
    as the test statistic. 
\end{enumerate}
the results were...

Figure~\ref{fig:gauss_mix_2} shows a histogram of the final minibatch sizes used
by the three methods in each iteration. Our method consumes significantly less
data; most minibatch sizes are smaller than 1000, and the average size is 210.
The other two methods occasionally need to consume a large proportion of the
entire data set of one million elements; the average minibatch sizes are 14862
and 18098 for Korattikara 2014 and Bardenet 2014, respectively. The average
minibatch sizes accurately predict the running times of these methods since
all methods have a running time proportional to the total data consumed, with
one caveat:

The method of~\cite{icml2014c1_bardenet14} requires a global bound on
the term $C_{\theta,\theta'}$ over the entire the data set at every
iteration in order to calculate the error bound (in equation (9) of
\cite{icml2014c1_bardenet14}).  This bound must be either computed
analytically using properties of the specific dataset, or it must be
checked at each sample if a black-box test is required. We found that
the sample code provided by the authors
of~\cite{icml2014c1_bardenet14} in fact computed this bound explicitly
for each sample generated. i.e. the implementation traversed the
entire dataset for each sample, providing no performance advantage
over the complete test. For this reason, we omitted the time to
compute $C_{\theta,\theta'}$ from our measurements of the running time
of the sample implementation of \cite{icml2014c1_bardenet14}, and
consider only the time to process the minibatches.

@@

>>>>

We start with a simple Gaussian mixture model, borrowing an experiment
from~\cite{langevin_2011}.  The parameter is 2-D, $\theta =
\langle \theta_1,\theta_2 \rangle$, and the parameter/data generation process is
%% \begin{equation}\label{eq:data_generation}
%%     (\theta_1,\theta_2) \sim \mathcal{N}((0,0), {\rm diag}(\sigma_1^2, \sigma_2^2));\quad \quad x_i \sim
%%     0.5 \cdot \mathcal{N}(\theta_1, \sigma_x^2) + 0.5 \cdot \mathcal{N}(\theta_1+\theta_2, \sigma_x^2).
%% \end{equation}
\begin{equation}\label{eq:data_generation}
    \begin{bmatrix} \theta_1 \\ \theta_2 \end{bmatrix} 
    \sim \mathcal{N}\left(\begin{bmatrix}0 \\ 0\end{bmatrix}, 
    \begin{bmatrix} \sigma_1^2 & 0 \\ 0 & \sigma_2^2\end{bmatrix} \right);\quad \quad x_i \sim
    0.5 \cdot \mathcal{N}(\theta_1, \sigma_x^2) + 0.5 \cdot \mathcal{N}(\theta_1+\theta_2, \sigma_x^2).
\end{equation}
We set $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$.  We fix $\theta =
\langle 0,1 \rangle$. The original paper sampled 100 data points from this
distribution, and estimated the posterior in the parameters. We are interested
in performance on larger problems and so sampled 1,000,000 points to form the
posterior of $p(\theta)\prod_{i=1}^{1,000,000}p(x_i\mid \theta)$ with the same
prior from Equation~\ref{eq:data_generation}. The larger number of samples
produces a much sharper posterior with two very narrow peaks.  Our goal is to
reproduce the original posterior, so we adjust the temperature to $T=10,000$.
Taking logs, we get the target as shown in the far left of
Figure~\ref{fig:gauss_mix_1}.

We run MCMC with our MH test using minibatch size $m=100$.  We also run this
using the methods from~\cite{cutting_mh_2014} (``Korattikara 2014'')
and~\cite{icml2014c1_bardenet14} (``Bardenet 2014''), also with $m=100$.
For~\cite{cutting_mh_2014}, we increment minibatches by 100 as recommended, and
for~\cite{icml2014c1_bardenet14}, we increase the minibatch size geometrically
with a ratio of $\gamma = 1.5$ as recommended. These tests use different
minibatch size updates because of their different approaches to repeated
testing. \cite{icml2014c1_bardenet14} uses straightforward discounting,
allocating a fraction of the test error $\delta$ to each minibatch test (call
these subtests). Its important to bound the number of subtests therefore, so
that the tolerable error per subtest is not too small.  \cite{cutting_mh_2014}
uses a more aggressive repeated testing approach, explicitly modeling
correlation between tests and allowing many more tests for a given $\delta$.

The tolerance for making a decision in~\cite{cutting_mh_2014} is
$\epsilon=0.005$, and the error bound control parameters
in~\cite{icml2014c1_bardenet14} are $p = 2$ and $\delta = 0.01$.  All methods
(including ours) use the same random walk proposer with covariance $\Sigma={\rm
diag}(0.3, 0.3)$.  Since the proposer generates a random walk, all shaping of
distribution of the samples is due to the MH test. All methods are run 5000
times to collect 5000 samples.

@@

Figure~\ref{fig:gauss_mix_1} shows scatter plots of the resulting $\theta$
samples for the three methods, with darker regions indicating a greater density
of points. There are no obvious differences in these distributions. To better
understand any differences, we measure the similarity between each set of
samples and the actual posterior.

To effect the comparison, we first discretize the posterior coordinates into
bins (non-overlapping ranges) with respect to the two components of the
parameter $\theta$.  The probability of a sample falling into one of these bins
is the integral of the true posterior probability over the area of that bin. Let
this probability be denoted $P_i$ where $i$ is the bin number. A single sample
from any of the MH methods should therefore be multinomial with distribution
$P$, and a set of $n$ (ideally independent) samples should follow a ${\rm
Multinomial}(P,n)$ distribution.  Since the ideal distribution is simple, we can
use it to measure the quality of the sample distributions rather than use
general purpose tests like KL-divergence or likelihood-ratio.  The latter in any
case are problematic with zero counts in some bins as we have here.

In practice, it is difficult to compute multinomial probabilities for large $n$,
in that case the per-bin distributions are well-approximated by Poisson
distributions with parameter $\lambda_i=P_i n$. Given a set of samples $\Theta =
\{\theta_1,\ldots,\theta_T\}$, let $c_j$ denote the number of individual samples
$\theta_i$ that fall in bin $j$. Then the log probability for the set is
approximated as
\begin{equation}\label{eq:log_prob_poissons}
    \log p(\Theta) \approx \sum_{j=1}^{N_{\rm bins}} c_j \log (n P_j) - n P_j - \log(\Gamma(c_j+1)).
\end{equation}

@@

While the Poisson results provide general guidance on the relative accuracy of
the models it is difficult interpret the quantitative scores. We therefore
perform significance tests to test the null hypothesis that the ideal
distribution really generated the samples.  We perform two tests:

\begin{enumerate}
    \item An exact test using simulation to generate an approximate distribution
    of scores given the null hypothesis.

    \item An approximate test using the chi-squared distribution as the test
    statistic. 
\end{enumerate}

The results were... {\color{blue} Daniel: IN PROGRESS}

Figure~\ref{fig:gauss_mix_2} shows a histogram of the final minibatch sizes used
by the three methods in each iteration. Our method consumes significantly less
data; most minibatch sizes are smaller than 1000, and the average size is 210.
The other methods occasionally need to consume a large proportion of the
entire data set of one million elements; the average minibatch sizes are 14862
and 18098 for Korattikara 2014 and Bardenet 2014, respectively. The average
minibatch sizes accurately predict the running times of these methods since
all methods have a running time proportional to the total data consumed, with
one caveat: the method of~\cite{icml2014c1_bardenet14} requires a global bound
on the term $C_{\theta,\theta'}$ over the entire the data set at every iteration
in order to calculate the error bound (in Equation~\ref{eq:bad_bound} of this
paper, i.e., Equation 9 of~\cite{icml2014c1_bardenet14}).  This bound must be
either computed analytically using properties of the specific dataset, or it
must be checked at each sample if a black-box test is required.  We found that
the sample code provided by the authors of~\cite{icml2014c1_bardenet14} in fact
computed this bound explicitly for each sample generated, i.e. the
implementation traversed the entire dataset for each sample, providing no
performance advantage over the complete test. For this reason, we omitted the
time to compute $C_{\theta,\theta'}$ from our measurements of the running time
of the sample implementation of \cite{icml2014c1_bardenet14}, and consider only
the time to process the minibatches.

@@

\begin{table}[t]
    \caption{{\color{blue} Daniel: TODO Poisson Probabilities, also we want
    ground truth right?}}
    \label{tab:poissons}
    \centering
    \begin{tabular}{l l l l l}
    \toprule
    Model & Ground Truth & Ours & Korattikara 2014 & Bardenet 2014 \\
    \midrule
    Poission Value & TODO & TODO & TODO & TODO \\
    \bottomrule
    \end{tabular}
\end{table}
